---
title: "Modifying `atable`"
author: "Alan Haynes"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Modifying `atable`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '')
library(atable)
```

Most things in `atable` are customizable. This vignette gives examples for some of them. See the `extending atable` vignette for adding methods for new classes. We will the `mtcars` datasets to demonstrate the concepts. 

```{r}
data(mtcars)
# factors
mtcars$am <- factor(mtcars$am, c(0, 1), c("Automatic", "Manual"))
mtcars$vs <- factor(mtcars$vs, c(0, 1), c("V-shaped", "straight"))
# ordered
mtcars$cyl <- ordered(mtcars$cyl)
# set format_to
atable_options(format_to = "md")
```
The `atable` default settings produce the following:
```{r, results='asis'}
knitr::kable(atable(vs + cyl + hp + disp ~ am, mtcars))
```


There are in general three approaches to modifying `atable`'s default settings:

1. pass a function to the `atable` function as an option. This affects only a single call to `atable`.
2. pass a function to `atable_options`. This affects all calls to `atable` during the session. These settings can be overriden by 1
3. replace a method in `atable`s namespace. This can be done with any package, but requires more code and is not easily reverted.

## Changing the statistical functions

In order to use any of these methods, we need suitable functions. Perhaps it is desirable to have t-tests and Kolmogorov-Smirnoff tests simultaneously. When defining test functions, the function has to have arguments `value`, `group` and `...` and return a named list.

```{r}
new_two_sample_htest_numeric <- function(value, group, ...){
  d <- data.frame(value = value, group = group)
  group_levels <- levels(group)
  x <- subset(d, group %in% group_levels[1], select = "value", drop = TRUE)
  y <- subset(d, group %in% group_levels[2], select = "value", drop = TRUE)
  ks_test_out <- stats::ks.test(x, y)
  t_test_out <- stats::t.test(x, y)
  out <- list(p_ks = ks_test_out$p.value,
              p_t = t_test_out$p.value)
  return(out)
}
```

Rather than mean and SD, maybe we want median and MAD, as well as the mean and SD. Statistics functions require arguments `x` and `...` and should return a named list. The class of the output should also be defined so that the appropriate formatting function can be selected.

```{r}
new_statistics_numeric <- function(x, ...){
  statistics_out <- list(Median = median(x, na.rm = TRUE),
                         MAD = mad(x, na.rm = TRUE),
                         Mean = mean(x, na.rm = TRUE),
                         SD = sd(x, na.rm = TRUE))
  class(statistics_out) <- c("statistics_numeric", class(statistics_out))
  # We will need this new class later to specify the format
  return(statistics_out)
}
```

We also need a function to format the statistics results (the default simply prints all elements of the statistics object, one after the other). Formatting functions require arguments `x` and `...` and should return a dataframe with variable tag (a factor) and value (a character). 

```{r}
new_format_statistics_numeric <- function(x, ...){
  Median_MAD <- paste(round(c(x$Median, x$MAD), digits = 1), collapse = "; ")
  Mean_SD <- paste(round(c(x$Mean, x$SD), digits = 1), collapse = "; ")
  levs <- c("Median; MAD", "Mean; SD")
  out <- data.frame(tag = factor(levs, 
                                 levels = levs),
                    # the factor needs levels for the non-alphabetical order
                    value = c(Median_MAD, Mean_SD),
                    stringsAsFactors = FALSE)
  return(out)
}
```


To use these three functions, we need to tell `atable` about them, using one of the three methods mentioned above. We will assign the testing function to `atable`s namespace. `two_sample_htest.numeric` will be used for two sample tests until R is restarted.

```{r}
utils::assignInNamespace(x = "two_sample_htest.numeric",
                         value = new_two_sample_htest_numeric,
                         ns = "atable")
```

`atable_options` can be used to specify a new function instead by referring to the thing to be replaced, in this case `statistics.numeric`. The `new_statistics_numeric` function will be used until e.g. R is restarted or `atable_options_reset` is used to restore the defaults, or it is replaced by something else.

```{r}
atable_options("statistics.numeric" = new_statistics_numeric)
```

The third option is to refer to the function in the `atable` call...

```{r}
knitr::kable(atable(hp + disp ~ am, mtcars, 
                    format_statistics.statistics_numeric = new_format_statistics_numeric))
```


Options that have been set via `atable_options` can be restored to there defaults via

```{r}
atable_options_reset()
```



## Labels 

`atable` can use Hmisc::label and Hmisc::unit, or an `alias` attribute to store labels. By default `labelled` objects (those with labels from Hmisc) are formatted with the label followed by the units in square brackets.

```{r}
require(Hmisc)
label(mtcars$hp) <- "Horse power"
units(mtcars$hp) <- "hp"
knitr::kable(atable(hp + disp ~ 1, mtcars))
```

Analagous to elsewhere, the `get_alias` methods are responsible for handling how the label is composed. If there is no `alias` attribute and the class is not `labelled`, then the variable name will be used. We can use round brackets instead of square ones by replacing the `get_alias.labelled` method...

```{r}
get_alias.labelled <- function(x, ...){
    out <- attr(x, "label", exact = TRUE)
    Units <- attr(x, "units", exact = TRUE)
    out = if(!is.null(Units)){
      paste0(out, " (", Units, ")")}else{out}
    return(out)
}
atable_options("get_alias.labelled" = get_alias.labelled)
knitr::kable(atable(hp + disp ~ 1, mtcars))
```

If we wanted to use a `label` attribute rather than `alias`, we could change the `get_alias.default` function
```{r}
attr(mtcars$disp, "label") <- "Displacement"
get_alias.default <- function(x, ...){
    attr(x, "label", exact = TRUE)
}
atable_options("get_alias.default" = get_alias.default)
knitr::kable(atable(hp + disp ~ 1, mtcars))
```





## P value formatting

By default, p values are formatted to 2 significant digits:

```{r}
atable_options("format_p_values")(0.12)
atable_options("format_p_values")(0.012)
atable_options("format_p_values")(0.0012)
atable_options("format_p_values")(0.0009)
```

This is easy to change by overwriting the function in `atable_options`. Here we report it to 3 decimal places.

```{r}
fn <- function(x){
  txt <- sprintf("%3.3f", x)
  if(x < 0.001) txt <- "<0.001"
  return(txt)
}
atable_options("format_p_values" = fn)
```

```{r}
atable_options("format_p_values")(0.12)
atable_options("format_p_values")(0.012)
atable_options("format_p_values")(0.0012)
atable_options("format_p_values")(0.0009)
```

The changes are then carried over to the table itself too. 

```{r}
knitr::kable(atable(vs + cyl + hp + disp ~ am, mtcars))
```

It should be noted that as there is no additional formatting function for the test results, so the default method is being used for the new. We would have to assign a class to the object returned from `new_two_sample_htest_numeric` and define a suitable test formatting function.

