---
title: "Modifying `atable`"
author: "Alan Haynes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying `atable`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '')
library(atable)
```

Most things in `atable` are customizable. This vignette gives examples for some of them. See the `extending atable` vignette for adding methods for new classes. We will the `mtcars` datasets to demonstrate the concepts. 

```{r}
data(mtcars)
# factors
mtcars$am <- factor(mtcars$am, c(0, 1), c("Automatic", "Manual"))
mtcars$vs <- factor(mtcars$vs, c(0, 1), c("V-shaped", "straight"))
# ordered
mtcars$cyl <- ordered(mtcars$cyl)
# set format_to
atable_options(format_to = "md")
```
The `atable` default settings produce the following:
```{r, results='asis'}
knitr::kable(atable(vs + cyl + hp + disp ~ am, mtcars))
```

There are in general three approaches to modifying `atable`'s default settings:

1. pass a function to the `atable` function as an option. This affects only a single call to `atable`.
2. pass a function to `atable_options`. This affects all calls to `atable` during the session. These settings can be overriden by 1
3. replace a method in `atable`s namespace. This can be done with any package, but requires more code and is not easily reverted.

In order to use any of these methods, we need suitable functions. Perhaps it is desirable to have t-tests and Kolmogorov-Smirnoff tests simultaneously. When defining test functions, the function has to have arguments `value`, `group` and `...` and return a named list.

```{r}
new_two_sample_htest_numeric <- function(value, group, ...){
  d <- data.frame(value = value, group = group)
  group_levels <- levels(group)
  x <- subset(d, group %in% group_levels[1], select = "value", drop = TRUE)
  y <- subset(d, group %in% group_levels[2], select = "value", drop = TRUE)
  ks_test_out <- stats::ks.test(x, y)
  t_test_out <- stats::t.test(x, y)
  out <- list(p_ks = ks_test_out$p.value,
              p_t = t_test_out$p.value)
  return(out)
}
```

Rather than mean and SD, maybe we want median and MAD, as well as the mean and SD. Statistics functions require arguments `x` and `...` and should return a named list. The class of the output should also be defined so that the appropriate formatting function can be selected.

```{r}
new_statistics_numeric <- function(x, ...){
  statistics_out <- list(Median = median(x, na.rm = TRUE),
                         MAD = mad(x, na.rm = TRUE),
                         Mean = mean(x, na.rm = TRUE),
                         SD = sd(x, na.rm = TRUE))
  class(statistics_out) <- c("statistics_numeric", class(statistics_out))
  # We will need this new class later to specify the format
}
```

We also need a function to format the statistics results (the default simply prints all elements of the statistics object, one after the other). Formatting functions require arguments `x` and `...` and should return a dataframe with variable tag (a factor) and value (a character). 

```{r}
new_format_statistics_numeric <- function(x, ...){
  Median_MAD <- paste(round(c(x$Median, x$MAD), digits = 1), collapse = "; ")
  Mean_SD <- paste(round(c(x$Mean, x$SD), digits = 1), collapse = "; ")
  levs <- c("Median; MAD", "Mean; SD")
  out <- data.frame(tag = factor(levs, 
                                 levels = levs),
                    # the factor needs levels for the non-alphabetical order
                    value = c(Median_MAD, Mean_SD),
                    stringsAsFactors = FALSE)
  return(out)
}
```

To use these three functions, we need to tell `atable` about them, using one of the three methods mentioned above. We will assign the testing function to `atable`s namespace.

```{r}
utils::assignInNamespace(x = "two_sample_htest.numeric",
                         value = new_two_sample_htest_numeric,
                         ns = "atable")
```












## P value formatting

By default, p values are formatted to 2 significant digits:

```{r}
atable_options("format_p_values")(0.12)
atable_options("format_p_values")(0.012)
atable_options("format_p_values")(0.0012)
atable_options("format_p_values")(0.0009)
```

This is easy to change by overwriting the function in `atable_options`. Here we report it to 3 decimal places.

```{r}
fn <- function(x){
  txt <- sprintf("%3.3f", x)
  if(x < 0.001) txt <- "<0.001"
  return(txt)
}
atable_options("format_p_values" = fn)
```

```{r}
atable_options("format_p_values")(0.12)
atable_options("format_p_values")(0.012)
atable_options("format_p_values")(0.0012)
atable_options("format_p_values")(0.0009)
```


